---
title: "Introduction to egvtools"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Introduction to egvtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(egvtools)
library(sf)
library(terra)
library(sfarrow)
library(landscapemetrics)
```

# Overview

`egvtools` provides tools for reproducible preparation of high-resolution
ecogeographical variables (EGVs) from heterogeneous spatial data. The package
is primarily intended for ecological applications in which environmental
information from multiple sources must be transformed into spatially aligned,
analysis-ready raster layers, for example as predictors in species distribution
models (SDMs).

Species distribution models relate observations of species occurrence or
abundance to environmental conditions and are widely used to describe
species-environment relationships and predict distributions across space and
time [@elith2009]. Consequently, the quality, ecological relevance, spatial
resolution, and spatial scale of environmental predictors are important parts
of the modelling workflow.

The main purpose of `egvtools` is not to replace general-purpose spatial
software. Instead, it provides wrappers and utilities for recurring operations
in large EGV-production workflows, including rasterization, grid
harmonization, aggregation, distance calculations, multi-scale summaries,
landscape metrics, and preparation of reusable spatial templates. The package
builds on established R spatial infrastructure, particularly `terra`, `sf`,
`sfarrow`, `exactextractr`, `landscapemetrics`, and `whitebox`
[@pebesma2018; @hesselbarth2019; @lindsay2016].

A particular emphasis is placed on reproducibility. Raster origin, resolution,
extent, coordinate reference system (CRS), grid-cell locations, and naming
conventions are kept consistent across processing steps. This is especially
important when large collections of predictors are produced independently and
later combined into a common modelling dataset.

The workflow implemented in `egvtools` builds on previous work developing
multi-scale EGVs for species distribution modelling in Latvia
[@avotins2022]. Although the examples in this vignette use Latvian template
grids, the underlying functions can also be applied to other study areas when
suitable reference grids and raster templates are provided.

The package was developed within the project HiQBioDiv: High-resolution
quantification of biodiversity for conservation and management, funded by the
Latvian Council of Science (Ref. No. VPP-VARAM-DABA-2024/1-0002).

The development version can be installed from GitHub with:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("aavotins/egvtools")
```

Spatial data processing terminology varies among disciplines and projects.
For clarity, `egvtools` uses the following operational terms throughout its
documentation. These describe stages of the EGV-production workflow rather
than formal or universal GIS data categories.

 - **Raw geodata** are source data obtained to describe environmental
conditions. They may be tables containing coordinates, raster datasets, or
vector datasets such as points, lines, and polygons. Raw geodata may already
have undergone basic preprocessing by their data provider, but they have not
yet been harmonized specifically for the EGV workflow.

 - **Geodata products** are intermediate datasets derived from one or more raw
geodata sources through substantial spatial processing. Such processing may
include overlays, reclassification, prioritization among overlapping
datasets, or combination of several data sources. In the workflows described
here, geodata products are commonly categorical raster layers with a defined
CRS, resolution, origin, and cell alignment.

Creating a geodata product can be important when overlapping source datasets
require explicit decisions about which spatial class takes precedence. For
example, a high-resolution raster cell cannot simultaneously represent both
open water and forest when a subsequent analysis requires an unambiguous
forest--water boundary.

 - **Input data or input layers** are high-resolution raster layers used
directly to calculate EGVs. Their spatial resolution is commonly finer than
the final EGV resolution. Working with harmonized raster inputs makes many
repeated operations considerably simpler and faster because decisions about
projection, alignment, overlaps, and class precedence have already been
resolved.
 - **Ecogeographical variables (EGVs)** are the analysis-ready environmental
variables produced by the workflow. They describe environmental conditions,
landscape composition, configuration, distance, diversity, or related
characteristics on a common spatial grid. EGVs can consequently be combined
directly with occurrence or other ecological data in statistical analyses,
including species distribution models. Multi-scale EGVs can additionally
represent environmental characteristics surrounding a focal grid cell at
several spatial scales [@avotins2022; @bradter2013].

In simplified form, the workflow can therefore be represented as:

**raw geodata -> geodata products -> harmonized input layers -> EGVs**

Not every dataset must pass through every stage. For example, a continuous
climate raster may be converted directly from raw geodata to an EGV, whereas
several overlapping land-cover datasets may first need to be combined into a
geodata product.


# Example data


Several small example datasets are included with `egvtools`. They are intended
to demonstrate the workflow without requiring the full national datasets used
in production analyses.

The examples below use harmonized raster and vector grids, grid centroids,
climate data, and Land Use Land Cover (LULC) information represented by
(CORINE Land Cover 2018)[https://land.copernicus.eu/en/products/corine-land-cover/clc2018].

CORINE Land Cover provides a harmonized European land-cover classification and
is used here as a readily interpretable categorical landscape dataset. The
example data are intended to demonstrate `egvtools`; users should select source
datasets with spatial and thematic resolution appropriate to their own
ecological questions.

```{r data}
data("clc18", package = "egvtools")
```

Raster example data distributed with the package are stored in `inst/extdata`
and can be accessed using `system.file()`.

```{r env-path}
climate_file <- system.file("extdata",
                        "Climate.tif",
                        package = "egvtools")

climate=terra::rast(climate_file)
terra::plot(climate)
```

The example data distributed with the package are deliberately small. The full
harmonization templates used for national processing can instead be downloaded
from Zenodo:

- (vector data)[https://doi.org/10.5281/zenodo.14237929];

- (raster data)[https://doi.org/10.5281/zenodo.14494873].

These templates define the spatial reference used throughout the workflow,
including CRS, raster origin, resolution, cell alignment, and identifiers used
for tiled processing.

```{r}
pkg_extdata <- system.file(
  "extdata",
  package = "egvtools"
)

grid_path100 <- file.path(
  pkg_extdata,
  "Templates",
  "TemplateGrids",
  "tikls100_sauzeme.parquet"
)

grid_path500 <- file.path(
  pkg_extdata,
  "Templates",
  "TemplateGrids",
  "tikls500_sauzeme.parquet"
)

template_path100 <- file.path(
  pkg_extdata,
  "Templates",
  "TemplateRasters",
  "LV100m_10km.tif"
)

template_path10 <- file.path(
  pkg_extdata,
  "Templates",
  "TemplateRasters",
  "LV10m_10km.tif"
)

template_path500 <- file.path(
  pkg_extdata,
  "Templates",
  "TemplateRasters",
  "LV500m_10km.tif"
)


work_dir <- file.path(tempdir(), "egvtools_vignette")
dir.create(work_dir, recursive = TRUE, showWarnings = FALSE)
```



# Reproducibility functions


Large spatial workflows are vulnerable to subtle inconsistencies among
datasets. Two rasters may have the same nominal resolution and CRS while still
having different origins or cell boundaries, making direct cell-by-cell
operations inappropriate.

For this reason, `egvtools` uses reusable vector and raster templates to define
the geometry of an analysis. The template download functions provide a
convenient way to recreate the same spatial framework on another computer or
in a new project directory.

### Downloading vector templates

`download_vector_templates()` downloads the reference grids and point layers
used for tiled and multi-scale calculations.

```{r download-vector-templates,eval=FALSE}
download_vector_templates(url="https://zenodo.org/api/records/22912421/files-archive",
  grid_dir = "../Templates/TemplateGrids",
  points_dir = "../Templates/TemplateGridPoints",
  gpkg_dir = "../Templates",
  overwrite = TRUE,
  quiet = FALSE)
```


### Downloading raster templates

`download_raster_templates()` retrieves the corresponding raster templates.
Together, the vector and raster templates allow independently generated EGVs
to retain exactly the same spatial geometry.

```{r download-raster-templates,eval=FALSE}
download_raster_templates(url="https://zenodo.org/api/records/22912630/files-archive",
  out_dir = "../Templates/TemplateRasters",
  overwrite = TRUE,
  quiet = FALSE)
```



# Preparing for core workflows

National-scale processing can involve millions of grid cells and very large
input datasets. `egvtools` therefore supports dividing calculations into
spatial tiles and reassembling the results afterwards.

### Tiling reference grids

`tile_vector_grid()` divides a reference grid according to an existing tile
identifier. This allows subsequent calculations to operate on manageable
subsets instead of repeatedly loading the complete national grid.

```{r tile-vector-grid}

tiles_grid <- file.path(
  work_dir,
  "TemplateGrids",
  "tiles"
)

dir.create(
  tiles_grid,
  recursive = TRUE,
  showWarnings = FALSE
)

tile_vector_grid(
  grid_path = grid_path100,
  out_dir = tiles_grid,
  tile_field = "tks50km",
  overwrite = TRUE,
  quiet = FALSE)
```

### Preparing multi-scale buffers

Ecological relationships are frequently scale-dependent. Environmental
conditions immediately surrounding an observation may describe local habitat,
whereas conditions measured over larger neighbourhoods can describe landscape
composition, habitat availability, fragmentation, or broader ecological
processes. Evaluating predictors at more than one spatial scale can therefore
be important in species distribution modelling [@bradter2013].

`tiled_buffers()` prepares reusable buffer geometries around grid-cell
centroids. Because these geometries are created once and subsequently reused
for many input variables, this can substantially reduce duplicated spatial
processing.

```{r tiled-buffers}
in_dir <- system.file(
  "extdata",
  "Templates",
  "TemplateGridPoints",
  package = "egvtools"
)

out_path_buff <- file.path(
  work_dir,
  "TemplateGridPoints",
  "tiles"
)

dir.create(
  out_path_buff,
  recursive = TRUE,
  showWarnings = FALSE
)


tiled_buffers(
  in_dir = in_dir,
  out_dir = out_path_buff,
  buffer_mode = "dense",
  radii_dense = c(500, 3000),
  split_field = "tks50km",
  n_workers = 2,
  future_max_mem_gb = 4,
  overwrite = TRUE,
  quiet = FALSE
)
```

With `buffer_mode = "dense"`, buffers are created around the full set of
reference points for each requested radius. The appropriate radii should be
chosen according to the ecological processes and taxa being studied rather
than interpreted as universally applicable spatial scales.

### Creating background rasters

Some transformations require an explicit value outside the features represented
by an input layer. For example, a binary habitat layer may contain `1` for the
target habitat and `NA` elsewhere, while an aggregation operation requires
non-target area to be represented by `0`.

`create_backgrounds()` creates raster backgrounds aligned with the reference
templates. These can subsequently be used to distinguish true zero values from
missing data.

```{r create-backgrounds}
in_dir <- system.file(
  "extdata",
  "Templates",
  "TemplateRasters",
  package = "egvtools"
)

out_path <- file.path(
  work_dir,
  "TemplateRasters"
)

dir.create(
  out_path,
  recursive = TRUE,
  showWarnings = FALSE
)


create_backgrounds(
  in_dir = in_dir,
  out_dir = out_path,
  background_value = 0,
  out_prefix="nulls_",
  overwrite = TRUE,
  terra_todisk = TRUE
)
```


# Core analysis functionality

The core functions in `egvtools` implement several common transformations from
source spatial information to analysis-ready EGVs. The examples below
illustrate a typical sequence rather than a mandatory pipeline. Individual
functions can also be used independently.



## Polygons to harmonized inputs

Many environmental datasets are supplied as polygons, whereas repeated
high-resolution calculations are often more efficient once categorical
information has been converted to a common raster grid.

`polygon2input()` rasterizes vector features using an existing raster template.
The template determines the CRS, resolution, extent, origin, and cell
locations, ensuring that the resulting input raster is compatible with other
layers in the workflow.

```{r clc18}

clc18$code=as.numeric(clc18$code_18)
polygon2input(vector_data = clc18,
              template_path = template_path10,
              out_path = work_dir,
              file_name = "clc_classes.tif",
              value_field = "code",
              plot_result = TRUE,
              plot_gaps = TRUE,
              overwrite = TRUE)
```

Here, CLC polygons are converted to a categorical raster at the resolution of
the high-resolution input template. The resulting raster is an intermediate
input layer rather than the final EGV.

Where several vector datasets overlap, class precedence should preferably be
resolved before or during this step. This ensures that subsequent landscape
boundaries represent ecological classes rather than inconsistencies among
source datasets.


## Inputs to harmonized EGVs

A common EGV is the proportion of a focal land-cover class within each analysis
cell. This can be calculated from a high-resolution binary input raster by
aggregating fine-resolution cells to the EGV grid.

```{r clc-rast}
classes <- file.path(
  work_dir,
  "clc_classes.tif"
)

clc_rast=terra::rast(classes)
forests=ifel(clc_rast>310&clc_rast<320,1,NA)
terra::plot(forests)
```

In this example, forest cells are selected from the categorical CLC raster.

```{r input2egv}

input2egv(input=forests,
          egv_template = template_path100,
          summary_function = "average",
          missing_job = "CoverInput",
          input_bg = 0,
          input_template = template_path10,
          check_alignment = TRUE,
          outlocation = work_dir,
          outfilename="Forest_prop.tif",
          layername="Forest_prop",
          plot_gaps = TRUE,
          plot_final = TRUE
          )
```

`input2egv()` summarizes the high-resolution input within cells of the EGV
template. With `summary_function = "average"` and a binary input consisting of
`1` for forest and `0` for non-forest, the resulting value represents the
proportion of each EGV cell occupied by forest.

The explicit background raster is important here because `NA` and zero have
different meanings: `NA` indicates unavailable information, whereas zero
represents a valid observation of no forest cover.

## Harmonizing coarse-resolution rasters

Not all environmental data originate at a finer resolution than the target EGV
grid. Climate and other continuous environmental datasets are frequently
available on substantially coarser grids.

`downscale2egv()` provides a standardized workflow for transferring such
continuous data to the EGV template, with optional gap filling and spatial
smoothing.

```{r downscale2egv}
df <- downscale2egv(
  template_path = template_path100,
  grid_path     = grid_path100,
  rawfile_path  = climate_file,
  out_path      = work_dir,
  file_name     = "climate_egv.tif",
  layer_name    = "climate",
  fill_gaps     = TRUE,
  smooth        = TRUE,
  smooth_radius_km = 10,
  plot_result   = TRUE
)
print(df)
```

The term downscaling here refers to producing a raster on the finer EGV grid;
it should not be interpreted as creating new independent environmental
information at that finer resolution. The effective information content
remains constrained by the resolution and methodology of the original
dataset. This distinction is particularly important when fine-resolution EGVs
are subsequently used in statistical models.

## Distance to an environmental feature

Distance variables provide another way to describe landscape structure. Rather
than measuring the amount of a land-cover class within a cell, they quantify
the spatial separation between locations and selected environmental features.

```{r distance2egv}
distance2egv(
  input         = clc_rast,
  template_egv  = template_path100,
  values_as_one = c("(310,320)"),
  outlocation   = work_dir,
  outfilename   = "dist_forest.tif",
  layername     = "dist_forest",
  use_whitebox  = TRUE,
  plot_result   = TRUE,
  plot_gaps     = TRUE,
  terra_todisk  = TRUE
)
```

In this example, cells representing forest are treated as the target object and
the output gives the distance to forest on the EGV grid. Such variables can be
useful when ecological responses are associated with proximity to habitat,
water, settlements, roads, or other spatial features.

When requested, `distance2egv()` can use WhiteboxTools as the computational
backend for distance calculations [@lindsay2016].

## Multi-scale zonal statistics

The value of an environmental predictor may depend on the spatial
neighbourhood over which it is measured. A local forest-cover value, for
example, describes conditions at the focal location, whereas forest cover
within 500 m or 3000 m describes increasingly broad landscape contexts.
Species can respond to environmental characteristics at several such scales
[@bradter2013].

`generalized_radius_function()` calculates summaries of one or more raster
layers within previously generated buffers and returns the results to the
common EGV grid.

```{r generalized-radius-function}
generalized_radius_function(
  kvadrati_path = tiles_grid,
  radii_path = out_path_buff,
  radii=c(500,3000),
  reference_grid_path = grid_path100,
  template_path = template_path100,
  input_layers = c(forests = file.path(work_dir,"/Forest_prop.tif")),
  output_dir = work_dir,
  layer_prefixes = c("ForestProp"),
  buffer_mode = "dense",
  n_workers = 2
)


terra::plot(terra::rast(file.path(work_dir,"/ForestProp_r500.tif")))
terra::plot(terra::rast(file.path(work_dir,"/ForestProp_r3000.tif")))
```

The two output layers therefore describe the same underlying variable at
different spatial scales. Keeping the scale in the output name is important
because these layers represent distinct candidate predictors even though they
originate from the same source raster.

The choice of radii should reflect the intended ecological interpretation.
Multiple scales are useful for evaluating scale dependence, but they can also
produce correlated predictors and should therefore be considered explicitly
during subsequent model selection.

## Landscape diversity

Landscape composition can also be summarized using metrics developed in
landscape ecology. `landscape_function()` provides a tiled interface to metrics
implemented by `landscapemetrics` [@hesselbarth2019].

```{r landscape-diversity}
res_tbl <- landscape_function(
  landscape      = clc_rast,
  zones          = grid_path500,
  id_field       = "rinda500",
  tile_field     = "tks50km",
  template       = template_path500,
  out_dir        = work_dir,
  out_filename   = "Landscape_diversity.tif",
  out_layername  = "Landscape_diversity",
  what           = "lsm_l_shdi",
  rasterize_engine = "fasterize",
  n_workers      = 2,
  fill_gaps      = TRUE,
  plot_gaps      = TRUE,
  plot_result    = TRUE
)
print(res_tbl)
terra::plot(terra::rast(res_tbl$output_path))
```

Here, `lsm_l_shdi` calculates Shannon's Diversity Index (SHDI) for the
categorical landscape within each analysis zone. The metric is

[
SHDI = -\sum_{i=1}^{m} p_i \ln(p_i),
]

where (p_i) is the proportion of the landscape occupied by class (i).
SHDI equals zero when only one landscape class is present and increases as the
number of classes and/or their evenness increases [@hesselbarth2019].

Consequently, this EGV represents compositional diversity rather than the area
of any particular habitat class.

## Edge lengths

Landscape configuration can be described not only by the amount of different
land-cover classes but also by their spatial arrangement. Edge metrics quantify
boundaries between neighbouring classes and can therefore describe aspects of
fragmentation, patch structure, or habitat interfaces [@hesselbarth2019].

For a binary forest/non-forest raster, total edge provides a simple measure of
the amount of boundary between the two classes within each analysis zone.


```{r edges}
forests2=terra::ifel(is.na(forests),0,1)

rez_edges <- landscape_function(
  landscape        = forests2,
  zones          = grid_path500,
  id_field       = "rinda500",
  tile_field     = "tks50km",
  template       = template_path500,
  out_dir          = work_dir,
  out_filename     = "edges_forests.tif",
  out_layername    = "edges_forests",
  what             = "lsm_l_te",
  lm_args          = list(count_boundary = FALSE),
  rasterize_engine = "fasterize",
  n_workers        = 2,
  future_max_size  = 4 * 1024^3,
 fill_gaps   = TRUE,
 plot_gaps   = FALSE,
 plot_result  = FALSE,
 keep_tiles = FALSE
)
rez_edges
edges=terra::rast(file.path(work_dir,"edges_forests.tif"))
plot(edges)
```

With `what = "lsm_l_te"`, `landscape_function()` calculates total edge length.
Setting `count_boundary = FALSE` prevents the outer boundary of the analysis
zone from being counted as landscape edge. This is generally desirable when
the ecological quantity of interest is the boundary between classes within the
zone rather than the arbitrary geometry of the sampling unit.

As with other landscape metrics, edge length is scale-dependent. Its value is
affected by both the resolution of the categorical input raster and the size
of the analysis zone. These should therefore be reported alongside the
resulting EGV.

# References
