---
title: "Introduction to egvtools"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to egvtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(egvtools)
library(sf)
library(terra)
library(sfarrow)
```

# Overview

`egvtools` provides a coherent set of wrappers and utilities that make large-scale 
EGV creation reproducible and pleasant on real datasets. The package leans on 
robust building blocks — `terra`, `sf`, `sfarrow`, `exactextractr`, and `whitebox` — and 
standardizes I/O, naming conventions, and multi-scale zonal statistics so your 
pipelines are repeatable across machines and projects.

The package was developed in project “HiQBioDiv: High-resolution quantification 
of biodiversity for conservation and management” funded by the Latvian Council 
of Science (Ref. No. VPP-VARAM-DABA-2024/1-0002).

The development version can be installed from GitHub with:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("aavotins/egvtools")
```

# Terminology

Athough all georeferenced data can be considered geodata, in this material we use the following terms in the order listed below in our workflows:

- **raw geodata** - considered as raw data obtained for a harmonised description of the environment. This may include tables with coordinates, raster or vector data. It can be anything that has been or can be used to create ecogeographical variables, with or without slight processing.

- **geodata product** - processed raw geodata that have undegone heavy modifications, e.g. spatial overlays and combinations of different sets of raw geodata, and are used as input data. In this document, geodata products are categorical raster layers that match the CRS and the pixel locations of input data. When split by categories, they become input data. The processing step of creating geodata products is necessary when decisions about the order of spatial overlays are important. For example, in a high-resolution pixel, there can only be water or forest, if the edge between water and forest need to be calculated.

- **input data or input layers** - very-high resolution (multiple times higher than that used for ecogeographical variables) raster data that are the direct input for the creation of most of the ecogeographical variables. The creation of such layers is particularly useful alongside geodata products, as dealing with border misalignment or decisions regarding the order of spatial o verlays, as well as simple geoprocessing, is much faster with raster data.

- **ecogeographical variables (EGVs)** - this is the final product of the workflow describing environment for statistical analysis (e.g. species distribution modelling). They are suitable also for publishing due to standadisation of the values. In other words, these are standardised landscape ecological variables in the form of high-resolution raster layers.


# Example data

Several example datasets are included in the package. The examples in this
vignette use harmonized raster and vector grids, grid centroids as point layers, 
climate and Land Use Land Cover (LULC) data represented by (Corine Land Cover (CLC) 
for the year 2018)[https://land.copernicus.eu/en/products/corine-land-cover/clc2018].

```{r data,eval=FALSE}
data("aves_small", package = "sdmhelpers")
data("bryophyta_one", package = "sdmhelpers")
data("habitat_inventories", package = "sdmhelpers")
data("points100", package = "sdmhelpers")
data("example_grid", package = "sdmhelpers")
example_grid=terra::unwrap(example_grid)
data("example_presences", package = "sdmhelpers")
data("example_background", package = "sdmhelpers")
data("example_independent_blocks", package = "sdmhelpers")
data("egv_names", package = "sdmhelpers")
```

The raster data are stored in `inst/extdata`.

```{r env-path,eval=FALSE}
egv_file <- system.file("extdata",
                        "example_egv.tif",
                        package = "sdmhelpers")

egv_file
```

Harmonization data can also be downloaded from Zenodo repository:

- (vector data)[https://doi.org/10.5281/zenodo.22912420];

- (raster data)[https://doi.org/10.5281/zenodo.22912629].


